---
title: "Exploring_tidyverse"
author: "Fran Camacho"
date: "2025-04-20"
output: html_document
---


## Exploring tidyverse

[Chapter 12 from the book "Machine Learning with R", by Brett Lantz].


```{r}

if (!require(tidyverse)) install.packages('tidyverse', dependencies = T)
library(tidyverse)

```

output:

[...]

tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
ℹ Use the conflicted package to force all conflicts to become errors


### Tidy table structures with tibbles

vignette("tibble")  ->  "Tibbles are a modern take on data frames. They keep the features that have stood the test of time, 
and drop the features that used to be convenient but are now frustrating."

```{r}
titanic_csv <- read.csv("./CSVs/titanic_train.csv")
titanic_tbl <- as_tibble(titanic_csv)
```

To show the "tibble":

```{r}
titanic_tbl
```
One simple distinction between tibbles and data frames is that a tibble never assumes
stringsAsFactors = TRUE, which was the default behavior in base R until relatively recently
with the release of R version 4.0.

Another distinction between tibbles and data frames is that, as long as the name is surrounded by the backtick (`) character, 
a tibble can use non-standard column names like `my var` that violate base R’s object naming rules.


### Reading rectangular files faster with readr and readxl

Nearly every chapter so far has used the read.csv() function to load data into R data frames.
Although we could convert these data frames in to tibbles, there is a faster and more direct path
to get data into the tibble format.

```{r}
#library(readr) # not necessary if tidyverse is already loaded
titanic_train_tbl <- read_csv("./CSVs/titanic_train.csv")       # read_csv, with "_"
titanic_train_tbl
```

For more information on the syntax, view the documentation using the vignette("readr") command.

The readxl package (https://readxl.tidyverse.org) provides a method to read data directly
from the Microsoft Excel spreadsheet format.

```{r}
library(readxl)
titanic_train_xls <- read_excel("./CSVs/titanic_train.xlsx")
titanic_train_xls
```

### Preparing and piping data with dplyr

The dplyr grammar of working with data becomes second nature after the initial learning curve
has been passed. There are five key verbs in the grammar, which perform many of the most com-
mon transformations to data tables. Beginning with a tibble, one may choose to:

•**filter()** rows of data by values of the columns
•**select()** columns of data by name
•**arrange()** rows of data by sorting the values
•**mutate()** columns into new columns by transforming the values
•**summarize()** rows of data by aggregating values into a summary


```{r}
titanic_csv |> filter(Sex == "female")
```

```{r}
titanic_tbl |> select(Name, Sex, Age)
```
```{r}
titanic_csv_women <- titanic_csv |> 
                      filter(Sex == "female") |> 
                      select(Name, Sex, Age) |> 
                      arrange(Age)
titanic_csv_women
```
When combined with the mutate() verb, we can perform complex data transformations with simpler, 
more readable code than in the base R language.


```{r}
titanic_train_tbl |>
              mutate(elderly = if_else(Age >= 65, 1, 0))
```

```{r}
titanic_train_tbl |>
            filter(Sex == "female") |>    # 314 rows
            mutate(
                  elderly = if_else(Age >= 65, 1, 0),
                  child = if_else(Age < 18, 1, 0)
            )
```

The remaining dplyr verb, **summarize()**, allows us to create aggregated or summarized metrics
by grouping rows in the tibbl.

```{r}

titanic_train_tbl |>
                group_by(Sex) |>
                summarize(survival_rate = mean(Survived))
```

Other more complicated example to compare children (less than 18 years old) to adults:

```{r}
titanic_train_tbl |>
            filter(!is.na(Age)) |>
            mutate(child = if_else(Age < 18, 1, 0)) |>
            group_by(child) |>
            summarize(survival_rate = mean(Survived))
```


To put together what we’ve learned so far and provide one more example using pipes, let’s build
a decision tree model of the Titanic dataset.

```{r}
if (!require(rpart)) install.packages('rpart', dependencies = T)
library(rpart)  # "recursive partitioning" package offers the most faithful implementation of regression trees
```

```{r}
m_titanic <- titanic_train_tbl |>
              filter(!is.na(Age)) |>
              mutate(AgeGroup = if_else(Age < 18, "Child", "Adult")) |>
              select(Survived, Pclass, Sex, AgeGroup) |>  
              rpart(formula = Survived ~ ., data = _)   
```

The **data = _** argument uses the underscore (_) as a placeholder to represent the data being fed to rpart() by the pipe.

[
It is important to note that the underscore placeholder character is new as of R version 4.2 and will not work in prior versions! 
In older code that uses the magrittr package, the dot character (.) was used as the placeholder.
]

```{r}
# more "standard/fran" way:
titanic_train_tbl_processed <- titanic_train_tbl |>
              filter(!is.na(Age)) |>
              mutate(AgeGroup = if_else(Age < 18, "Child", "Adult")) |>
              select(Survived, Pclass, Sex, AgeGroup)  
              
m_titanic2 <- rpart(formula = Survived ~ ., data = titanic_train_tbl_processed)     
```


For fun, we can visualize the resulting decision tree, which shows that women and children are
more likely to survive than adults, men, and those in third passenger class:

```{r}
if (!require(rpart.plot)) install.packages('rpart.plot', dependencies = T)
library(rpart.plot)  # "recursive partitioning" package offers the most faithful implementation of regression trees
```

```{r}
rpart.plot(m_titanic)
rpart.plot(m_titanic2)
```



### Transforming text with stringr

The free resource R for Data Science has a tutorial that introduces the package’s complete set of capabilities, at 

https://r4ds.hadley.nz/strings.html ,

but here, we’ll examine some of the aspects most relevant to feature engineering.

All stringr functions begin with the prefix str_, and a few relevant
examples are as follows:

•**str_detect()** determines whether a search term is found in a string
•**str_sub()** slices a string by position and returns a substring
•**str_extract()** searches for a string and returns the matching pattern
•**str_replace()** replaces characters in a string with something else

we’ll begin by examining the Cabin feature to determine whether certain
rooms on the Titanic are linked to greater survival:

```{r}
titanic_train_tbl_with_CabinCode <- titanic_train_tbl |>
                                mutate(CabinCode = str_sub(Cabin, start = 1, end = 1)) # |> 
                                # filter(!is.na(CabinCode))
```

Show records:

```{r}
head(titanic_train_tbl_with_CabinCode)
```
To confirm that the cabin code is meaningful, we can use the table() function to see a clear 
relationship between it and the passenger class. The useNA parameter is set to "ifany" to display
the NA values caused by missing cabin codes for some passengers:

```{r}
table(titanic_train_tbl_with_CabinCode$Pclass, titanic_train_tbl_with_CabinCode$CabinCode, useNA = "ifany")
```

The NA values appear to be more common in the lower ticket classes, so it seems plausible that
cheaper fares may have not received a cabin code.

We can also plot the survival probability by cabin code by piping the file into a ggplot() function:

```{r}
library(ggplot2)

titanic_train_tbl_with_CabinCode |> ggplot() +
  geom_bar(aes(x = CabinCode, y = Survived),
  stat = "summary", fun = "mean") +
  ggtitle("Titanic Survival Rate by Cabin Code")
```
The resulting figure shows that even within the first-class cabin types (codes A, B, and C) there
are differences in survival rate.
Additionally, the passengers without a cabin code are the least likely to survive.

Without processing the Cabin text data first, a learning algorithm would be unable to use the
feature as the codes are unique to each cabin. Yet by applying a simple text transformation, we’ve
decoded the cabin codes into something that can be used to improve the model’s survival predictions.



```{r}
titanic_train_tbl_with_CabinCode_and_Title <- titanic_train_tbl |>
                                              mutate(Title = str_extract(Name, ", [A-z]+\\.")) |> 
                                              mutate(Title = str_replace_all(Title, "[, \\.]", ""))

head(titanic_train_tbl_with_CabinCode_and_Title$Title)
```

```{r}
table(titanic_train_tbl_with_CabinCode_and_Title$Title)
```

Given the small counts for some of these titles and salutations, it may make sense to group them
together. To this end, we can use dplyr’s recode() function to change the categories.

```{r}
titanic_train_tbl_with_CabinCode_Title_and_TitleGroup <- 
  titanic_train_tbl_with_CabinCode_and_Title |>
    mutate(TitleGroup = recode(Title,
                                "Mr" = "Mr", "Mrs" = "Mrs", "Master" = "Master",
                                "Miss" = "Miss", "Ms" = "Miss", "Mlle" = "Miss", "Mme" = "Miss",
                                .missing = "Other",
                                .default = "Other"
                              )
          )
```

```{r}
table(titanic_train_tbl_with_CabinCode_Title_and_TitleGroup$TitleGroup)
```

We can also see that the title is meaningful by examining a plot of survival rates by title:

```{r}
titanic_train_tbl_with_CabinCode_Title_and_TitleGroup |> ggplot() +
  geom_bar(aes(x = TitleGroup, y = Survived), stat = "summary", fun = "mean") +
  ggtitle("Titanic Survival Rate by Salutation")
```

The creation of CabinCode and TitleGroup features exemplifies the feature engineering technique
of finding hidden information in text data.



### Cleaning dates with lubridate

The lubridate package (https://lubridate.tidyverse.org) is an important tool for working
with date and time data. 

```{r}
d1 <- mdy(c("October 25, 2013", "9/20/2013"))
d1
```

```{r}
d2 <- dmy(c("25 October 2013", "22.11.19"))
d2
```


```{r}
ymd("2013-10-25")
```

Let’s create a similar object for each of the three previous editions of this book:

```{r}
MLwR_1stEd <- mdy("October 25, 2013")
MLwR_2ndEd <- mdy("July 31, 2015")
MLwR_3rdEd <- mdy("April 15, 2019")
```

We can do simple math to compute the difference between two dates:

```{r}
MLwR_2ndEd - MLwR_1stEd
```

```{r}
MLwR_3rdEd - MLwR_2ndEd
```

!!!

In years?

```{r}
as.duration(MLwR_2ndEd - MLwR_1stEd)

as.duration(MLwR_3rdEd - MLwR_2ndEd)
```

To obtain only years:

```{r}
dyears()  # it gives seconds in a year
```

So now:

```{r}
as.duration(MLwR_2ndEd - MLwR_1stEd) / dyears()

as.duration(MLwR_3rdEd - MLwR_2ndEd) / dyears()
```

Other way, using **"time_length()"** function:

```{r}
time_length(MLwR_2ndEd - MLwR_1stEd, unit = "years")

time_length(MLwR_3rdEd - MLwR_2ndEd, unit = "months")
```


```{r}
A_LA_BASTILLA <- mdy("July 14, 1789") 
time_length(mdy("April 21 2025") - A_LA_BASTILLA, unit = "years")
```

```{r}
time_length(mdy("July 13 2025") - A_LA_BASTILLA, unit = "years")
```

```{r}
interval(A_LA_BASTILLA, mdy("July 20 2023")) / years()  # ACHTUNG! interval() uses start, end syntax !!!
```

To return only the integer component of the age use int. div. **%/%**:

```{r}
interval(A_LA_BASTILLA, mdy("July 20 2023")) %/% years()  
```


Generalizing this work, we can create a function to compute the calendar-based age for a given
date of birth as of today:

```{r}
age <- function(birthdate) {
  birthdate %--% today() %/% years()    # %--% shorthand for interval construction operator
}
```


Tests:

```{r}
  age(mdy("January 20, 1952")) 
  age(mdy("April 20, 1950")) 
```


