---
title: "Missing_data"
author: "Fran Camacho"
date: "2025-04-23"
output: word_document
---

# Dealing with missing data

[Chapter 12 from the book "Machine Learning with R", by Brett Lantz].

## Understanding types of missing data

1. Data **missing completely at random (MCAR)** is independent of the other features and
its own value; in other words, it would not be possible to predict whether any particular
value is missing.

2. Data **missing at random (MAR)** may depend on other features but not on the underlying
value, which means that certain, predictable rows are more likely than others to contain
missing values.

3. Data **missing not at random (MNAR)** is missing due to a reason related to the missing
value itself. This data is in essence censored from the dataset for some reason impossible
to discern from the other features in the dataset.


## Performing missing value imputation



```{r}
if (!require(tidyverse)) install.packages('tidyverse', dependencies = T)
library(tidyverse)
```


```{r}
titanic_train <- titanic_train |>
    mutate(
      Cabin = if_else(is.na(Cabin), "X", Cabin),
      Embarked = if_else(is.na(Embarked), "Unknown", Embarked)
    )
```



### 1) Simple imputation with missing value indicators

In addition to imputing a value in place of the NA, it is especially important to create a 
**missing value indicator (MVI)**, which is a binary variable that indicates whether the 
feature value was imputed.

```{r}
titanic_train <- titanic_train |>
            mutate(
                  Age_MVI = if_else(is.na(Age), 1, 0),
                  Age = if_else(is.na(Age), mean(Age, na.rm = TRUE), Age)
            )
```


**Both the feature that has been imputed and the MVI should be included as predictors in the**
**machine learning model.**

[..] if a value is missing, there is probably an explanation for it. 
For example, in the Titanic dataset, perhaps the missing age implies something 
about the passenger’s social status or family background.


### 2) Missing value patterns

A missing value pattern is essentially a character string composed of a series of MVIs, 
with each character in the string representing a feature with missing values.

[..] . For example, the first row would be represented by the string '11100000', 
which indicates that the first three features for this loan applicant were missing. 
The second applicant, who had no missing data, would be represented by '00000000', 
hile the second and third would be represented by '00000111' and '01011101', respectively.

Although missing value patterns can be extremely powerful predictors, they are not without some
challenges. First and foremost, in a dataset containing k features, there are 2 exp k potential values of
a missing value pattern. A dataset with just 10 features may have as many as 1,024 levels of the
MVP predictor, while a dataset with 25 features would have over 33 million potential levels







