---
title: "Sparse_data"
author: "Fran Camacho"
date: "2025-04-23"
output: word_document
---

# Dealing with sparse data

[Chapter 12 from the book "Machine Learning with R", by Brett Lantz].


```{r}
if (!require(tidyverse)) install.packages('tidyverse', dependencies = T)
library(tidyverse)
```


## Example – Remapping sparse categorical data

```{r}
titanic_train <- read_csv("./CSVs/titanic_train.csv") |>
                  mutate(Title = str_extract(Name, ", [A-z]+\\.")) |>
                  mutate(Title = str_replace_all(Title, "[, \\.]", ""))

table(titanic_train$Title, useNA = "ifany")
```


```{r}
titanic_train <- titanic_train |>
  mutate(TitleGroup = fct_collapse(Title,
          Mr = "Mr",
          Mrs = "Mrs",
          Master = "Master",
          Miss = c("Miss", "Mlle", "Mme", "Ms"),
          Noble = c("Don", "Sir", "Jonkheer", "Lady"),
          Military = c("Capt", "Col", "Major"),
          Doctor = "Dr",
          Clergy = "Rev",
          other_level = "Other")
        ) |>
        mutate(TitleGroup = fct_na_value_to_level(TitleGroup, level = "Unknown"))
```

Examining the new categorization, we see that the 17 original categories have been reduced to 9:

```{r}
table(titanic_train$TitleGroup)
```

The **fct_count()** function provides a sorted list of the feature levels and their proportions of the overall total:

```{r}
fct_count(titanic_train$Title, sort = TRUE, prop = TRUE)
```
With **fct_lump()_X** set of functions we can take the top three levels and treat everything else as other:

```{r}
table(fct_lump_n(titanic_train$Title, n = 3))
```

Alternatively, we can lump together all levels with less than one percent of the observations:

```{r}
table(fct_lump_prop(titanic_train$Title, prop = 0.01))
```

Lastly, we might choose to lump together all levels with less than five observations:

```{r}
table(fct_lump_min(titanic_train$Title, min = 5))
```



## Example – Binning sparse numeric data

```{r}
titanic_train <- read_csv("./CSVs/titanic_train.csv")
```

Let's apply discretization techniques to the fare values

```{r}
head(titanic_train$Fare, 10)
```

```{r}
summary(titanic_train$Fare)
```

```{r}
titanic_train <- titanic_train |> 
  mutate(fare_firstclass = if_else(Fare >= 31, 1, 0, missing = 0)
)
```

This reduces a feature with nearly 250 distinct values into a new feature with only two:

```{r}
table(titanic_train$fare_firstclass)
```

**if_else** is not appropiated when there are more than 2 options.
tidyverse function called **case_when()** allows the construction of a more complex series of checks to determine the result:

```{r}
titanic_train <- titanic_train |>
  mutate(
    fare_class = case_when(
                          Fare >= 31 ~ "1st Class",
                          Fare >= 15 ~ "2nd Class",
                          TRUE ~ "3rd Class"
                )
        )
```

Now we have

```{r}
table(titanic_train$fare_class)
```


The **cut()** function.
It'is included in base R and provides a simple method for creating a factor from
a numeric vector. 

The breaks parameter specifies the cut points for the numeric range, shown as follows for a three-level factor 
that matches the previous discretization. 

The right = FALSE parameter indicates that the levels should not include the rightmost, or highest, value and the
Inf break point indicates that the final category can span the range of values from 31 to infinity.

```{r}
table(cut(titanic_train$Fare, breaks = c(0, 15, 31, Inf), right = FALSE))
```

By default, cut() sets labels for factors that indicate the range of values.

cut combined with seq:

```{r}
table(cut(titanic_train$Fare, right = FALSE, breaks = seq(from = 0, to = 550, by = 50)))
```

As an alternative to having equally sized intervals, we can construct bins with an equal number of
examples (using **quantile()**:

```{r}
table(cut(titanic_train$Fare, right = FALSE,
              breaks = quantile(titanic_train$Fare, probs = seq(0, 1, 0.20))))
```

Note that the bins do not contain exactly the same number of examples due to the presence of ties.

The tidyverse also includes a function for creating quantile-based groups, which may be easier to
use in some cases. This ntile() function divides the data into n groups of equal size. For example,
it can create five groups as follows:

```{r}
table(ntile(titanic_train$Fare, n = 5))
```

Because the function assigns the groups numeric labels, it is important to convert the resulting
vector to a factor. This can be done directly in a mutate() statement:

```{r}
titanic_train <- titanic_train |>
            mutate(fare_level = factor(ntile(Fare, n = 11)))
```

The resulting feature is

```{r}
table(titanic_train$fare_level)
```

Although the level still has numeric labels, because the feature has been coded as a factor, it will
still be treated as categorical by most R functions.





