---
title: "Tema10_roc_curve"
author: "Fran Camacho"
date: "2025-03-31"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ROC curve


[From the book "Machine Learning with R", by Brett Lantz].


package pROC:
  
https://web.expasy.org/pROC/    


```{r}
install.packages("pROC")    
library(pROC)
```


Import the CSV file with the data already prepared:

```{r}
sms_results <- read.csv(file.path("Chapter10", "sms_results.csv"), stringsAsFactors=TRUE)
```

To create visualizations with pROC, two vectors of data are needed. 
The first must contain the # estimated probability of the positive class and 
the second must contain the predicted class values.

```{r}
sms_roc <- roc(sms_results$actual_type, sms_results$prob_spam)
```

Graphic:

```{r}  
plot(sms_roc, main = "ROC curve for SMS spam filter",
     Col = "blue", lwd = 2, grid = TRUE, legacy.axes = TRUE)
```


### Compare this naive bayes model with another (i.e. a knn model)

```{r}
# import the CSV file for this other model
sms_results_knn <- read.csv(file.path("Chapter10", "sms_results_knn.csv"), stringsAsFactors=TRUE)

sms_roc_knn <- roc(sms_results$actual_type, sms_results_knn$p_spam)
```


Plot the graphic:

```{r}
plot(sms_roc, main = "ROC curve for SMS spam filter",
     Col = "blue", lwd = 2, grid = TRUE, legacy.axes = TRUE)

plot(sms_roc_knn, col = "red", lwd = 2, add = TRUE)
```

The curve for k-NN is consistently lower, suggesting that it is a consistently worse model than the Naive Bayes approach

### AUC

```{r}
auc(sms_roc)      # -> Area under the curve: 0.9836

auc(sms_roc_knn)  # -> Area under the curve: 0.8942
```


The AUC for the Naive Bayes SMS classifier is 0.98, which is extremely high and substantially
better than the k-NN classifier’s AUC of 0.89.