---
title: "Web_scraping_RSelenium_periodicos_online"
author: "Fran Camacho"
date: "2025-05-28"
output: word_document
---



Importar librerías

```{r}
if (!require(RSelenium)) install.packages('RSelenium', dependencies = T)
library(RSelenium) 

if (!require(rvest)) install.packages('rvest', dependencies = T)
library(rvest) 

if (!require(tidyverse)) install.packages('tidyverse', dependencies = T)
library(tidyverse) 

if (!require(wdman)) install.packages('wdman', dependencies = T)
library(wdman)

if (!require(stringr)) install.packages('stringr', dependencies = T) #regex
library(stringr)

if (!require(httr)) install.packages('httr', dependencies = T)
library(httr)

if (!require(xml2)) install.packages('xml2', dependencies = T)
library(xml2) 
```

Ojo:
✖ dplyr::filter()         masks stats::filter()
✖ readr::guess_encoding() masks rvest::guess_encoding()
✖ dplyr::lag()            masks stats::lag()




```{r}
# Start Selenium server:

# Linux console:  
# xvfb-run java -Dwebdriver.chrome.driver=/usr/local/bin/chromedriver -jar /usr/local/bin/selenium-server-standalone.jar

# Here:
#system("java -Dwebdriver.chrome.driver=/usr/local/bin/chromedriver -jar /usr/local/bin/selenium-server-standalone.jar &")

# Best way:
# run selenium_server$process as done below
```


### Arrancar servidor Selenium:

```{r}
# Start the Selenium server with the specified driver and capabilities

# selenium_server <- selenium(retcommand = TRUE, verbose = TRUE, check = FALSE) #debug

#binman::list_versions("chromedriver") # list chromedriver versions

extraCap <-  list("goog:chromeOptions" = list(
                      "args" = list("--headless")  # Optional: run in headless mode
                    )
                 )
                              
selenium_server <- selenium(port = 4444L, 
                            verbose = FALSE, 
                            extraCapabilities = extraCap
                           )
```

```{r}
#selenium_server
selenium_server$process # to show the running status and the pid
```


```{r}
# to stop the selenium server
selenium_server$stop()
```

```{r}
selenium_server$error()
print("-------------------")
selenium_server$log()
```

### Listados de páginas:


```{r}
# 1st link list

link01 <- "https://www.elmundo.es/television/momentvs/2025/05/22/682eb61ffdddff620c8b457f.html"    #81
summary01 <- "Felipe_González_Hormiguero"

link02 <- "https://www.elmundo.es/deportes/2025/05/26/68344b9921efa09a428b4583.html"      #248
summary02 <- "Euskadi_vs_España_pelota_vasca"

link03 <- "https://www.elmundo.es/espana/2025/05/26/6834b7e5e4d4d8664a8b4584.html"    #311
summary03 <- "Leire_Díez_UCO"

link04 <- "https://www.elmundo.es/espana/2025/05/29/683820d3e85eceba4a8b45b0.html"    #481
summary04 <- "Novio_Ayuso_procesado"

link05 <- "https://www.elmundo.es/espana/2025/05/30/68397ab9fc6c8361558b456d.html"    #361
summary05 <- "Periodistas_ElPais_se_ratifican_Supremo"

#1482 comments

links_vector <- c(link01,link02,link03,link04,link05)
summaries_vector <- c(summary01,summary02,summary03,summary04,summary05)
```


```{r}
# 2nd link list

link01 <- "https://www.elmundo.es/television/2025/05/30/6839d219e85eceb30a8b45b2.html"   #99
summary01 <- "Familia_Tele_cancelada_parcialmente"

link02 <- "https://www.elmundo.es/andalucia/2025/05/30/683877e5e4d4d89d1f8b4614.html"    #342
summary02 <- "UCO_85_enchufados"

link03 <- "https://www.elmundo.es/economia/dinero-inversion/2025/05/30/683991d0fc6c83734d8b4572.html"    #98
summary03 <- "IBEX35_mejor_mes"

link04 <- "https://www.elmundo.es/espana/2025/05/30/6839b825e4d4d898398b459e.html"    #171 
summary04 <- "Feijoo_pancartero"

link05 <- "https://www.elmundo.es/andalucia/2025/05/30/683990dde85eceb12d8b4593.html"   #92
summary05 <- "Juez_Sevilla_PSOE"

#802

links_vector <- c(link01,link02,link03,link04,link05)
summaries_vector <- c(summary01,summary02,summary03,summary04,summary05)
```


```{r}
# 3rd link list

link01 <- "https://www.elmundo.es/espana/2025/06/04/6840086321efa044528b4573.html"     #142
summary01 <- "Aldama_irrumpe_conferencia_Leire_Diez"

link02 <- "https://www.elmundo.es/espana/2025/06/04/683fea3021efa081758b45b7.html"     #432
summary02 <- "Leire_Diez_comparecencia_sin_preguntas"

link03 <- "https://www.elmundo.es/internacional/2025/06/03/683f1d9ae85eced8208b45a6.html"   #91
summary03 <- "España_niega_5_porciento_OTAN"

link04 <- "https://www.elmundo.es/internacional/2025/06/03/683ea3bde85ece3c108b458f.html"  #81
summary04 <- "Ultra_Geert_Wilders_abandona_gobierno_PPBB"

link05 <- "https://www.elmundo.es/espana/2025/06/04/683f3458e9cf4ad9038b45b6.html"      #81
summary05 <- "Gobierno_asume_falta_plazas_universidad_publica"

#827

links_vector <- c(link01,link02,link03,link04,link05)
summaries_vector <- c(summary01,summary02,summary03,summary04,summary05)
```


Total de mensajes obtenidos: 3111


Agrupamos los listados en dataframes:

```{r}
links_df <- data.frame(summary = summaries_vector, link = links_vector)
```

```{r}
links_df
```


### Procesar una página:


Función que recibe el nombre resumido y el enlace de la noticia:

```{r}
#Function to obtain the comments of the readers of a piece of news

read_page_el_mundo_function <- function(link_info) {
  
  remDr <- remoteDriver(port = 4444L, browserName = "chrome")
  remDr$open()

  link_info <- data.frame(link_info) 
    
  info <- link_info$summary
  print(info)
  page_link <- link_info$link
  print(page_link)
  
  remDr$navigate(page_link)  
  Sys.sleep(1)
                 
  try({                                               
    button_privacy <- remDr$findElement(using = "xpath", value = '/html/body/div[1]/div/div/div/div/div/div[2]/div[1]/p/div[2]/button[1]')
    #button_privacy <- remDr$findElement(using = "id", value = "ue-accept-notice-button")
    button_privacy$clickElement()  
  })
  
  button_show_comments <- remDr$findElement(using = "xpath", value = '/html/body/main[1]/div[5]/div/div/div[1]/aside[2]/button')
  button_show_comments$clickElement()
        
  Sys.sleep(1)
    

  comments_block <- remDr$findElement(using = "xpath", "/html/body/main[1]/div[7]")  
                                                            
  num_comments_html <- remDr$findElement(using = "class name", value = "js-numComments")
  number_of_posts <- as.integer(str_extract(num_comments_html$getElementAttribute("innerHTML"), "(?<=<strong>)\\d+(?=</strong>)"))
  print(paste("number of posts: ", number_of_posts))

  num_clicks <- round(number_of_posts / 10, 0)

  for (i in 1:num_clicks+1) {    
  
    comments_block$sendKeysToElement(list(key = "page_down"))
    comments_block$sendKeysToElement(list(key = "page_down"))
    comments_block$sendKeysToElement(list(key = "page_down"))
    
    button_show_more_comments <- remDr$findElement(using = "xpath", value = '/html/body/main[1]/div[7]/section/div/section/div[2]/ul/li/a')
    button_show_more_comments$clickElement()
        
    # wait
    Sys.sleep(0.1)
  }

  comments_block$sendKeysToElement(list(key = "page_down"))
  comments_block$sendKeysToElement(list(key = "page_down"))
  comments_block$sendKeysToElement(list(key = "page_down"))
  comments_block$sendKeysToElement(list(key = "page_down"))
  
  # Read the page source after clicking the button
  page_source <- remDr$getPageSource()[[1]]
  html_content <- read_html(page_source)

  # Extract all comments
  article_nodes <- html_content |>  html_elements("article")

  num_comments <- length(article_nodes)
  print(paste("number of comments: ",length(article_nodes)))

  article_nodes <- article_nodes[3:num_comments]

  #build lines author | comment:
  comments <- article_nodes |> 
    map(function(x) {
      divs <- html_nodes(x, "div")
      author <- divs[1] |>  html_text2()
      text <- divs[2] |>  html_text2()
      tibble(author = author, text = text)
    }) |> 
    bind_rows() |> 
    mutate(comments = paste(author, text, sep = "|")) |> 
    pull(comments)
  
  
  comments <- comments[1:(number_of_posts+20)]
  #comments

  file_name <- paste(info, "txt", sep = ".")
  #file_name <- gsub(" ","",file_name)
  #create file with the comments
  cat(comments, file=file_name, sep="\n")

  remDr$close()
}

```


```{r}
info <- links_df[3,]$summary

file_name <- paste(info, "txt",sep = ".")

print(file_name)

file_name <- gsub(" ","",file_name)
print(file_name)

```



Leer una página:

```{r}

read_page_el_mundo_function(links_df[3,])

```


Leer todas las páginas:


```{r}
#open remote driver  
#remDr <- remoteDriver(port = 4444L, browserName = "chrome")
#remDr$open()


# IT DOESN'T WORK!
#apply(links_df, 1, read_page_el_mundo_function)

# IT DOESN'T WORK!
#for (page in links_df) {
#  print(page)
#  read_page_el_mundo_function(page)
#}

read_page_el_mundo_function(links_df[1,])
read_page_el_mundo_function(links_df[2,])
read_page_el_mundo_function(links_df[3,])
read_page_el_mundo_function(links_df[4,])
read_page_el_mundo_function(links_df[5,])

#close remote driver
#remDr$close()
```


Cerrar el driver

```{r}
remDr$close()
```

```{r}
# stop the server
selenium_server$stop()
```









